---
title: "kNN imputation"
author: "eh"
date: "11/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r reading data}
ins_train_df <- read.csv("https://raw.githubusercontent.com/johnm1990/msds-621/main/insurance_training_data.csv")
ins_eval_df <- read.csv("https://raw.githubusercontent.com/johnm1990/msds-621/main/insurance-evaluation-data.csv")
```

# What do zeros and blanks and NA's have to do with each other?

```{r}
sum(ins_train_df$HOME_VAL == "$0")
sum(ins_train_df$HOME_VAL == '')
```

```{r}
sum(ins_train_df$INCOME == "$0")
sum(ins_train_df$INCOME == '')
```

```{r}
sum(!is.na(ins_train_df$YOJ) & ins_train_df$YOJ == 0.0)
sum(is.na(ins_train_df$YOJ))
```

```{r}
sum(ins_train_df$JOB == '')
```

```{r}
sum(ins_train_df$BLUEBOOK == "$0")
sum(ins_train_df$BLUEBOOK == '')
```

```{r}
sum(ins_train_df$OLDCLAIM == "$0")
sum(ins_train_df$OLDCLAIM == '')
```

```{r}
sum((ins_train_df$JOB == '') & (is.na(ins_train_df$YOJ)))
```

```{r}
sum((ins_train_df$JOB == '') & (!is.na(ins_train_df$YOJ) & ins_train_df$YOJ == 0.0))
```

**Only 31 of 526 blank fields for JOB align with NA or 0.0 for YOJ**

```{r}
sum((ins_train_df$INCOME == '') & (ins_train_df$JOB == ''))
```

```{r}
sum((ins_train_df$INCOME == "$0") & (ins_train_df$JOB == ''))
```

```{r}
sum((ins_train_df$INCOME == "$0") & (is.na(ins_train_df$YOJ)))
```

```{r}
sum((ins_train_df$INCOME == '') & (is.na(ins_train_df$YOJ)))
```

```{r}
sum((ins_train_df$INCOME == '') & (!is.na(ins_train_df$YOJ) & ins_train_df$YOJ == 0.0))
```

```{r}
sum((ins_train_df$INCOME == '$0') & (!is.na(ins_train_df$YOJ) & ins_train_df$YOJ == 0.0))
```

So most INCOME == '$0' is YOJ == 0.0, and half of them are home makers, the other half students.

```{r}
unemp = ins_train_df[((ins_train_df$INCOME == '$0') & (!is.na(ins_train_df$YOJ) & ins_train_df$YOJ == 0.0)),]
head(unemp)
mean(unemp$JOB=='Home Maker' | unemp$JOB=='Student')
```

```{r}
yojna = ins_train_df[is.na(ins_train_df$YOJ),]
mean(yojna$HOME_VAL=='')
mean(yojna$INCOME=='')
mean(yojna$JOB=='')
```

### Parse numbers from strings  

```{r}
dollars <- function(n){
  as.integer(str_remove_all(n, '[,$]')) 
}
dollars(ins_train_df$INCOME[1:11])
```

```{r}
yojna$INCOME = dollars(yojna$INCOME)
hist(yojna$INCOME)
```

NA's in YOJ are not a stand-in for unemployment.

Blanks in HOME_VAL and INCOME and JOB maybe similarly shouldn't be treated as 0's or unemployed or whatever, but rather NA's.  
Even if HOME_VAL == "" meant that someone rents instead of owns, setting HOME_VAL to 0 would create a false ordering with other home values.

```{r}
ins_train_df$INCOME = dollars(ins_train_df$INCOME)
ins_eval_df$INCOME = dollars(ins_eval_df$INCOME)
ins_train_df$HOME_VAL = dollars(ins_train_df$HOME_VAL)
ins_eval_df$HOME_VAL = dollars(ins_eval_df$HOME_VAL)
ins_train_df$JOB[ins_train_df$JOB==''] = NA
ins_eval_df$JOB[ins_eval_df$JOB==''] = NA
```

```{r}
table(ins_train_df$JOB)
sum(is.na(ins_train_df$JOB))
```

```{r}
hist(ins_train_df$INCOME)
```

```{r}
hist(ins_train_df$HOME_VAL)
```

```{r}
ins_train_df$OLDCLAIM = dollars(ins_train_df$OLDCLAIM)
ins_eval_df$OLDCLAIM = dollars(ins_eval_df$OLDCLAIM)
ins_train_df$BLUEBOOK = dollars(ins_train_df$BLUEBOOK)
ins_eval_df$BLUEBOOK = dollars(ins_eval_df$BLUEBOOK)
```

#### Turn boolean strings into booleans

```{r}
ins_train_df$REVOKED = tolower(ins_train_df$REVOKED)=='yes'
ins_eval_df$REVOKED = tolower(ins_eval_df$REVOKED)=='yes'
ins_train_df$PARENT1 = tolower(ins_train_df$PARENT1)=='yes'
ins_eval_df$PARENT1 = tolower(ins_eval_df$PARENT1)=='yes'
ins_train_df$RED_CAR = tolower(ins_train_df$RED_CAR)=='yes'
ins_eval_df$RED_CAR = tolower(ins_eval_df$RED_CAR)=='yes'
ins_train_df$MSTATUS = tolower(ins_train_df$MSTATUS)=='yes'
ins_eval_df$MSTATUS = tolower(ins_eval_df$MSTATUS)=='yes'
```

#### Turn unordered categorical strings into unordered factors

```{r}
ins_eval_df$SEX <- factor(ins_eval_df$SEX)
ins_train_df$SEX <- factor(ins_train_df$SEX)
ins_eval_df$JOB <- factor(ins_eval_df$JOB)
ins_train_df$JOB <- factor(ins_train_df$JOB)
ins_eval_df$CAR_USE <- factor(ins_eval_df$CAR_USE)
ins_train_df$CAR_USE <- factor(ins_train_df$CAR_USE)
ins_eval_df$CAR_TYPE <- factor(ins_eval_df$CAR_TYPE)
ins_train_df$CAR_TYPE <- factor(ins_train_df$CAR_TYPE)
ins_eval_df$URBANICITY <- factor(ins_eval_df$URBANICITY)
ins_train_df$URBANICITY <- factor(ins_train_df$URBANICITY)
```

#### Turn ordered categorical strings into ordered factors

```{r}
unique(ins_train_df$EDUCATION)
```

```{r}
ins_eval_df$EDUCATION <- factor(ins_eval_df$EDUCATION,
                                   levels = c("<High School", "z_High School",
                                               "Bachelors", "Masters", "PhD"),
                                    ordered = T)
ins_train_df$EDUCATION <- factor(ins_train_df$EDUCATION,
                                   levels = c("<High School", "z_High School",
                                               "Bachelors", "Masters", "PhD"),
                                    ordered = T)
```

```{r}
summary(ins_train_df)
```


```{r}
summary(ins_train_df)
```
# KNN-Impute NA's

```{r}
library(VIM)
```
First, mark the values as NA's, in case the knn-imputation doesn't work

```{r}
ins_train_df['NA_YOJ'] = is.na(ins_train_df$YOJ)
ins_eval_df['NA_YOJ'] = is.na(ins_eval_df$YOJ)
ins_train_df['NA_CAR_AGE'] = is.na(ins_train_df$CAR_AGE)
ins_eval_df['NA_CAR_AGE'] = is.na(ins_eval_df$CAR_AGE)
ins_train_df['NA_HOME_VAL'] = is.na(ins_train_df$HOME_VAL)
ins_eval_df['NA_HOME_VAL'] = is.na(ins_eval_df$HOME_VAL)
ins_train_df['NA_INCOME'] = is.na(ins_train_df$INCOME)
ins_eval_df['NA_INCOME'] = is.na(ins_eval_df$INCOME)
ins_train_df['NA_JOB'] = is.na(ins_train_df$JOB)
ins_eval_df['NA_JOB'] = is.na(ins_eval_df$JOB)
```

### The below kNN is an updated version of what was saved to github in the "write_csv" chunk following it.  Different attempts to change kNN hyperparameters provided only worse results.  So best to use what's already pushed to github.

```{r}
ins_train_df = kNN(ins_train_df, variable = c('YOJ', 'CAR_AGE', 'HOME_VAL',
                                              'INCOME', 'JOB'),
                dist_var = names(ins_eval_df[4:31]), k=5, weights = 'auto', imp_var=F,
                addRF=T, useImputedDist = F)
ins_eval_df = kNN(ins_eval_df, variable = c('YOJ', 'CAR_AGE', 'HOME_VAL',
                                              'INCOME', 'JOB'),
                dist_var = names(ins_eval_df[4:31]), k=5, weights = 'auto', imp_var = F,
                addRF=T, useImputedDist = F)
```

```{r}
summary(ins_train_df)
```