---
title: "eh621FP"
author: "Ethan Haley"
date: "11/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(dplyr)
```

```{r WHO}
WHO_URL <- "https://raw.githubusercontent.com/douglasbarley/DATA621/main/Final%20Project/Life%20Expectancy%20Data.csv"
WHO <- read_csv(WHO_URL )
names(WHO)[1] <- "CountryName"
head(WHO)
```

```{r WB_Pop}
WB_Pop_URL <- "https://raw.githubusercontent.com/douglasbarley/DATA621/main/Final%20Project/Population_by_Country.csv"
WB_Pop <- read_csv(WB_Pop_URL) %>%
  subset(select = c("CountryName","CountryCode"))
head(WB_Pop)
```


### World Bank (WB) Country Region metadata

```{r WB_Meta}
WB_Meta_URL <- "https://raw.githubusercontent.com/douglasbarley/DATA621/main/Final%20Project/Country_Region_Metadata.csv"
WB_Meta <- read_csv(WB_Meta_URL) %>%
  subset(select = c("CountryCode","Region","IncomeGroup"))
head(WB_Meta)
```

### Join WB data tables

```{r WB_data}
WB_data <- inner_join(WB_Pop,WB_Meta, by = c("CountryCode" = "CountryCode"), copy = FALSE, keep = FALSE)
head(WB_data, 10)
```
### Join WHO data with WB metadata for each country

```{r WHO_with_Region}
WHO_with_Region <- left_join(WHO,WB_data, by = c("CountryName" = "CountryName"), copy = FALSE, keep = FALSE)

head(WHO_with_Region, 10)

summary(WHO_with_Region)

```
## EDA 

```{r}

WHO_Final <- WHO_with_Region

# Dropping NA
WHO_Final <- WHO_Final %>%
  drop_na() 

# renaming columns 
WHO_Final <- WHO_Final %>%
  rename(LifeExpectancy = "Life expectancy", AdultMortality = "Adult Mortality", InfantDeaths = "infant deaths"
         , IncomeComposition ="Income composition of resources")

summary(WHO_Final)

```    

### Add columns for other life expectancy data that doesn't have inconsistencies

```{r}
lifexpURL <- 'https://raw.githubusercontent.com/ebhtra/msds-621/main/FinalProject/life_exp_breakdown.csv'
lifexp = read_csv(lifexpURL)
head(lifexp)
```

Pivot to combine all data for one country/one year into one wider row

```{r}
lifexp = pivot_wider(lifexp, names_from = c('Indicator'), values_from = c('Male','Female','Both_Sexes'))
lifexp = lifexp %>% fill(everything(), .direction = 'up')
lifexp = lifexp[seq(1, 1464, 2),]
tail(lifexp)
```
```{r}
# rename columns and drop first one
lifexp <- lifexp %>% select(-c(1)) %>%
  rename(MLife60 = "Male_Life expectancy at age 60 (years)",
         MLife0  = "Male_Life expectancy at birth (years)",
         FLife60 = "Female_Life expectancy at age 60 (years)",
         FLife0  = "Female_Life expectancy at birth (years)",
         Life60 = "Both_Sexes_Life expectancy at age 60 (years)",
         Life0  = "Both_Sexes_Life expectancy at birth (years)")
head(lifexp, 2)
```






```{r}
WHO = WHO[order(WHO$CountryName, WHO$Year),]
for (co in unique(WHO$CountryName)){
  coDF = WHO[WHO$CountryName==co,]
  plot(coDF$Year, coDF$`Life expectancy`, type='l', main = co, xlab = 'year', ylab = 'life expectancy')
}
```

```{r}
WHO[WHO$CountryName=="Cook Islands",]
```
Remove the few countries whose data are incomplete by year 

```{r}
cTable = table(WHO$CountryName)
incompletes = c()
for (co in unique(WHO$CountryName)){
  if (cTable[co] < 16) {incompletes = c(incompletes, co)}
}
incompletes
```

```{r}
WHO = WHO[!WHO$CountryName %in% incompletes,]
dim(WHO)
```

```{r}
govtExp = read_csv("~/CUNY_MSDS/DATA621/FinalProject/healthExpenditure.csv")
head(govtExp)
```
```{r}
dim(govtExp)
```

```{r}
table(govtExp$SpatialDimValueCode)
```

```{r}
govtExp = govtExp[order(govtExp$Location, govtExp$Period),]
head(govtExp)
```
```{r}
govtExp = subset(govtExp, select = c(SpatialDimValueCode, Location, Period, FactValueNumeric))
head(govtExp)
```
```{r}
summary(govtExp)
```

```{r}
print("Countries that make up the 168 missing values for govt expenditure %: ")
table(govtExp[is.na(govtExp$FactValueNumeric), 'Location'])
```


```{r}
WHO[WHO$Measles==max(WHO$Measles),]
```
Make the measles a rate instead of a number  

```{r}
stillBirths = read_csv('~/CUNY_MSDS/DATA621/FinalProject/stillBirthData.csv')
head(stillBirths)
```

Stillbirths per 1000 total births

```{r}
stillBirths = stillBirths[order(stillBirths$Location, stillBirths$Period),]
stillBirths = subset(stillBirths, select = c(SpatialDimValueCode, Location, Period, FactValueNumeric))
head(stillBirths)
```

```{r}
summary(stillBirths)
```


```{r}
WHO[WHO$`HIV/AIDS`==max(WHO$`HIV/AIDS`),]
```
### Try for better population numbers  

```{r}
library(rvest)
temp <- read_html('https://en.wikipedia.org/wiki/List_of_countries_by_past_and_projected_future_population')
pops <- temp %>%
  html_nodes("table") %>%
  html_table(fill = TRUE)
```

```{r}
pops
```

```{r}
pop = pops[[2]]
pop = pop %>% 
  select(c('Country (or dependent territory)', '2000', '2005', '2010', '2015')) %>%
  rename(Country = 'Country (or dependent territory)') 
pop
```
### Make values numeric and multiply by 1000 to match our WHO d.f.


```{r}
numb = function(x){as.integer(str_remove_all(x, ",")) * 1000}
pop[, 2:5] = sapply(pop[, 2:5], numb)
pop
```

### See if country names are compatible, for merging  

```{r}
for (n in unique(WHO_with_Region$CountryName)){
  if (! n %in% unique(pop$Country)){
    print(n)
  }
}
```

```{r}
unique(WHO_with_Region$CountryName)
```
```{r}
pop$Country[25] = "Bolivia (Plurinational State of)"
pop$Country[30] = "Brunei Darussalam"
pop$Country[99] = "Côte d'Ivoire"
pop$Country[37] = "Cabo Verde"
pop$Country[164] = "Congo"
pop$Country[51] = "Czechia"
pop$Country[147] = "Democratic People's Republic of Korea"
pop$Country[93] = "Iran (Islamic Republic of)"
pop$Country[110] = "Lao People's Democratic Republic"
pop$Country[66] = "Micronesia (Federated States of)"
pop$Country[190] = "Republic of Korea"
pop$Country[130] = "Republic of Moldova"
pop$Country[166] = "Russian Federation"
pop$Country[177] = "Sao Tome and Principe"
pop$Country[63] = "Swaziland"
pop$Country[198] = "Syrian Arab Republic"
pop$Country[148] = "The former Yugoslav republic of Macedonia"
pop$Country[215] = "United Kingdom of Great Britain and Northern Ireland"
pop$Country[201] = "United Republic of Tanzania"
pop$Country[216] = "United States of America"
pop$Country[221] = "Venezuela (Bolivarian Republic of)"
pop$Country[222] = "Viet Nam"
```

```{r}
for (n in unique(WHO_with_Region$CountryName)){
  if (! n %in% unique(pop$Country)){
    print(n)
  }
}
```

Pivot population values for different years into their own rows  

```{r}
poplong = pivot_longer(pop, cols = c('2000', '2005', '2010', '2015'),
                       names_to = 'Year', values_to = 'Pop')
poplong$Year = as.numeric(poplong$Year)
poplong
```
#### Merge into main d.f.  

```{r}
WHO_plus_pops = left_join(WHO_with_Region, poplong, by = c("CountryName"="Country", "Year"="Year"))
WHO_plus_pops
```

### Check for govt expenditure % if country codes match main DF, for merging purposes  

```{r}
for (n in unique(WHO_plus_pops$CountryCode)){
  if (! n %in% unique(govtExp$SpatialDimValueCode)){
    print(n)
  }
}
```

Reminder:  
```{r}
head(govtExp)
```
```{r}
merger = govtExp %>%
  rename(CountryCode = "SpatialDimValueCode",
         Year = "Period", PctExp = "FactValueNumeric") %>%
  select(-c(2))
head(merger)
```
```{r}
WHO_plus_healthExp = left_join(WHO_plus_pops, merger, by = c("CountryCode", "Year"))
head(WHO_plus_healthExp)
```
#### Repeat procedure with stillbirths data  
reminder:  

```{r}
head(stillBirths,3)
```
```{r}
for (n in unique(WHO_plus_healthExp$CountryCode)){
  if (! n %in% unique(stillBirths$SpatialDimValueCode)){
    print(n)
  }
}
```

```{r}
merger = stillBirths %>%
  rename(CountryCode = "SpatialDimValueCode",
         Year = "Period", StillBirthRate = "FactValueNumeric") %>%
  select(-c(2))
head(merger)
```

```{r}
WHO_plus_stillBirths = left_join(WHO_plus_healthExp, merger, by = c("CountryCode", "Year"))
head(WHO_plus_stillBirths)
```
### Now add the life expectancy columns  

```{r}
head(lifexp, 2)
```


```{r}
for (n in unique(WHO_plus_stillBirths$CountryCode)){
  if (! n %in% unique(lifexp$CCode)){
    print(n)
  }
}
```

```{r}
merger = lifexp %>%
  rename(CountryCode = "CCode") %>%
  select(c(2,3,5:11))
head(merger)
```
```{r}
WHO_plus_lifeExp = left_join(WHO_plus_stillBirths, merger, by = c("CountryCode", "Year"))
head(WHO_plus_lifeExp)
```
Rename some cols

```{r}
WHO_plus_lifeExp <- WHO_plus_lifeExp %>%
  rename(LifeExpectancy = "Life expectancy", AdultMortality = "Adult Mortality", 
         InfantDeaths = "infant deaths", IncomeComposition ="Income composition of resources")
names(WHO_plus_lifeExp)
```

#### Rearrange columns so we can compare values better  

```{r}
WHO_full = WHO_plus_lifeExp %>%
  select(c(1,2,4,35,34,33,32,31,30,8,27,18,26,7,21,22,25,3,17,28,6,12,5,13,15,16,9,10,14,19,20,11,24,29,23))
WHO_full[111:115,]
```

#### Remove the few countries whose data are incomplete by year 

```{r}
cTable = table(WHO_full$CountryName)
incompletes = c()
for (co in unique(WHO_full$CountryName)){
  if (cTable[co] < 16) {incompletes = c(incompletes, co)}
}
incompletes
```

```{r}
WHO_full = WHO_full[!WHO_full$CountryName %in% incompletes,]
dim(WHO_full)
```

```{r}
write_csv(WHO_full, "WHO_full.csv")
```





